### 2017-12-05 paper/sexSelectionSpacing-ver1.tex
Edits of intro and empirical strategy

What has not previously been appreciated is that the introduction 
of prenatal sex determination fundamentally changed the relationship 
between son preference and birth spacing.
Couples with stronger son preferences are more likely to use prenatal
sex determination and sex-selective abortions, but each sex-selective 
abortion substantially increases the duration from last birth. 
Each abortion increases spacing between births by approximately a year.
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of a spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which I set to 6 
months \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.%
\footnote{
The waiting time to conception varies by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}

As a result, we now have a situation where families with the \emph{strongest} son 
preference may show the \emph{longest} spacing after the birth of a daughter,
because of their use of sex-selective abortions.
But, to further complicate matters, we may still observe short spacing 
after the births of daughters as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

Spacing, by itself, can therefore no longer be used to measure son preference.
Understanding birth spacing remains, however, an important undertaking.
First, spacing is still important and useful in understanding son preference, 
if combined with the likelihood of observing a boy or a girl.
% Short spacing after a girl is indicative of son preference if observed
% births are close to the natural sex ratio, whereas long spacing is indicative
% of son preference if observed births are male biased.
Second, if spacing affect health outcomes for mother and/or children, it is 
important to understand what drives changes in spacing.
The better health outcomes for girls in the presence of sex selection may,
for example, be an unintended side-effect of the longer spacing that arise
from sex-selective abortions rather than be driven by a higher value of
girls resulting from lower supply \citep{Lin2014,Hu2015}.
Third, spacing may be an important factor in parents' decisions, either
for preference or economic reasons, such as labor force participation.
This is especially important here because even parents with strong
son preference may reverse their decision to use prenatal sex determination, 
and carry the next pregnancy to term whether male or female, as the duration 
from last birth becomes sufficiently sufficiently long.%
\footnote{
More generally, we know less about what determines spacing behavior in 
developing countries than in developed countries.
With increasing numbers of women entering the labor force in developing
countries, understanding how timing decisions are made will be important
for the design of suitable policies \citep{Portner2018}.
}

In this paper, I introduce and apply a novel method that directly incorporates 
the effects of sex-selective abortions on the likelihood of a son being born 
\emph{and} the duration between births.
The method can be used to analyse both situations with and without prenatal
sex selection.
Furthermore, my proposed method allow for the duration since the last 
birth to affect the decision on sex selection.
By examining whether sex selection decisions change with spacing, we can 
draw a more nuanced picture of the degree of son preference.

>>>>

What has not previously been appreciated is that the introduction 
of prenatal sex determination fundamentally changed the relationship 
between son preference and birth spacing.
Couples, who used to have short birth spacing because of son preference,
now have access to prenatal sex determination and sex-selective abortions,
but each sex-selective abortion they have substantially increases the space
between births. 
Birth spacing increases by approximately a year per abortion.
The increase consists of three parts.
First, starting from the time of the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of a spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which I set to 6 
months \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.%
\footnote{
The waiting time to conception varies by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}

As a result, we now have a situation where families with the \emph{strongest} son 
preference may show the \emph{longest} spacing after the birth of a daughter,
because they use sex-selective abortions.
But, to further complicate matters, we may still observe short spacing 
after the births of daughters as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

Spacing, by itself, can therefore no longer be used to measure son preference.
However, understanding birth spacing remains an important undertaking.
First, spacing is still useful in understanding son preference, 
if combined with the likelihood of observing a boy or a girl.
% Short spacing after a girl is indicative of son preference if observed
% births are close to the natural sex ratio, whereas long spacing is indicative
% of son preference if observed births are male biased.
Second, if spacing affects health outcomes for mother and/or children, it is 
important to understand what drives changes in spacing.
The better health outcomes for girls in the presence of sex selection may,
for example, be an unintended side-effect of the longer spacing that arise
from sex-selective abortions, rather than, as often assumed, be driven by a 
higher value placed on girls because of their lower number \citep{Lin2014,Hu2015}.
Third, spacing may be an important factor in parents' decisions, either
for preference or economic reasons.
This is especially important here, because even parents with strong
son preference may reverse their decision to use prenatal sex determination, 
and carry the next pregnancy to term whether male or female, as the duration 
from the prior birth becomes sufficiently sufficiently long.
Finally, we know less about what determines spacing behavior in 
developing countries than in developed countries.
With increasing numbers of women entering the labor force in developing
countries, understanding how timing decisions are made will be important
for the design of suitable policies \citep{Portner2018}.

In this paper, I introduce and apply a novel empirical method that directly incorporates 
the effects of sex-selective abortions on the likelihood of a son being born 
\emph{and} the duration between births.
The method can be used to analyse both situations with and without prenatal
sex selection.
Furthermore, my proposed method allow for the duration since the prior 
birth to affect the decision on sex selection.
By examining whether sex selection decisions change with spacing, we can 
draw a more nuanced picture of the degree of son preference.
